Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 2.dev

### 2024-09-24 main.tex
Do not use a hyphen in balance-assist.

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add a
  controlled steering motor to a consumer electric bicycle that stabilizes it
  at low speeds. To test the motor's assistance during falls, we apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill with the balance-assist system activated and deactivated. The
  probability of not recovering from a handlebar perturbation decreases when
  the balance-assist is activated at a travel speed of 6~\si{\kph}. Use of a
  balance-assist system in real world bicycling will reduce the number of falls
  that occur near riders' control authority limits.
}

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll rate~\footnote{Roll angular rate is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle at lower speeds. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the high speed self-stability
already does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance-assisting bicycle, we subject
the rider to perturbations at the handlebar, which can relatively easily cause
a rider to fall.

In this paper, we test whether a steer motor controlled bicycle, that is stable
in a large low speed range, is beneficial in helping to prevent the rider from
falling. We test this by applying varying magnitude mechanical perturbations to
the handlebars while the rider is cycling on a treadmill. We assess the
difference in the rider's probability of falling with the balance-assist system
on and off.

@@

Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single-actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to effectively change
the dynamics, up to some limits, of the human-controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance-assist system but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

@@

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability for a low speed range. We use this simple
feedback principle as the basis for our balance-assist controller.

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
(VR IMU BN0086, Sparkfun, Niwot, USA). The balance-assist control algorithm is
implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at at
least 100 Hz.

@@

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is determined
at random. If the perturbation results in a fall, the estimate of the
perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
perturbation did not result in a fall. Five new possible perturbation forces
are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance-assist system is on or off
during the first or second set of perturbations for each participant.

@@

We record the order number of each perturbation, i.e. first, second, third,
\ldots, to measure how many perturbations a rider is subjected to before the
current pertubration. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance-assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test our hypothesis that the balance-assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for the
statistical model variables.
%
\begin{table}
  \centering
  \caption{Independent and dependent variables used in the statistical model.}
  \label{tab:stat-model-variables}
  \begin{tabular}{llll}
    \toprule
    Variable & Causality & Units & Description \\
    \midrule
    \(L\)        & independent & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(\delta_0\) & independent & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\)   & independent & \si{\degree} & roll angle at start of perturbation \\
    \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
    \(c\)        & independent & boolean & balance-assist: off 0 or on 1 \\
    \(j\)        & independent & integer & order number of perturbation \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosen as all data from an individual
subject because we are only interested in the association between the state of
the balance-assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance-assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain $g=10$ trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance-assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are shown in Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    6~\si{\kph} and gain $g=10$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
angular impulse and perturbation order are statistically significant predictors
with angular impulse being the dominant effect. Larger angular impulse
increases the probability to fall and enduring more perturbations decreases the
probability to fall. Unlike the 6~\si{\kph} trials, the balance-assist state is
not a significant predictor. At both speeds the angular impulse has about twice
the effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    10~\si{\kph} and gain $g=8$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance-assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}

Turning the balance-assist system on significantly \((p<0.05)\) reduces the
odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance-assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probability from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance-assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance-assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewness of the probability curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance-assist system turned on, but the effect is not
significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance-assist on
    (solid) or off (dashed) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance-assist and the
    angular impulse. The abscissa is standard deviation around the mean (all
    perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
We have shown that at a 6~\si{\kph} riding speed the addition of balance-assist
control reduces the chance that a rider will fall when perturbed around the
limits of their control authority. But this effect diminishes just below
significance at the higher speed scenario of 10~\si{\kph}. We were only able to
test these two speed-gain scenarios for mostly homogeneous sets of riders
within the resources of this research project, but additional experimental work
could help understand more completely the range and limits of the positive
effect of the balance system. For example, it is possible that simply
increasing the controller gain at 10~\si{\kph} also results in a significant
positive effect. A longitudinal study of normal use of the balance-assist
bicycle compared to a control group could provide the strongest evidence of any
benefit we have seen in this more narrow scenario.

>>>>

\abstract{
  Uncontrolled bicycles are generally unstable at low speeds. We add a
  controlled steering motor to a consumer electric bicycle that stabilizes it
  at low speeds. To test the motor's assistance during falls, we apply varying
  magnitude external handlebar perturbations to the bicycle while ridden on a
  treadmill with the balance assist system activated and deactivated. The
  probability of not recovering from a handlebar perturbation decreases when
  the balance assist is activated at a travel speed of 6~\si{\kph}. Use of a
  balance assist system in real world bicycling will reduce the number of falls
  that occur near riders' control authority limits.
}

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles themselves are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of the necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll rate~\footnote{Roll angular rate is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle at lower speeds. If automatic control of steering can
stabilize a bicycle, it may reduce the control required from the rider to
successfully manage balancing tasks much like the high speed self-stability
already does. We have developed a balance assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can relatively easily cause
a rider to fall.

In this paper, we test whether a steer motor controlled bicycle, that is stable
in a large low speed range, is beneficial in helping to prevent the rider from
falling. We test this by applying varying magnitude mechanical perturbations to
the handlebars while the rider is cycling on a treadmill. We assess the
difference in the rider's probability of falling with the balance assist system
on and off.

@@

Many more automatically balanced single track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in balancing and possibly reduce single-actor falls. Most
of these robotic bicycle and motorcycle designers did not intend for a human
rider to also control the stabilized vehicle. Nevertheless, an automatically
stabilized bicycle can be controlled by a human rider if the motor controlled
steer torque and the rider applied steer torque act on the steer in parallel.
The effect of this automatic control gives the ability to effectively change
the dynamics, up to some limits, of the human-controlled plant. In our prior
study~\citep{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions due to the balance assist system but contrarily researchers with a
similar vehicle recently showed rider dissatisfaction with the
stabilization~\cite{Hanakam2023}.

@@

The state matrix \(\mathbf{A}\) being a function of the equilibrium speed \(v\)
means \(k_{\dot{\phi}}\) can be selected such that the eigenvalues of
\(\left(\mathbf{A} - \vec{B}_\delta\left[0 \quad k_{\dot{\phi}} \quad 0 \quad
0\right] \right)\) have negative real parts for \(v_{min} < v <
v_\textrm{capsize}\) where \(v_{min}\) is the lowest stable speed given
\(k_{\dot{\phi}}\) and \(v_\textrm{capsize}\) is the speed at which the
uncontrolled bicycle's capsize mode goes unstable. With gain scheduling with
respect to \(v\), the speed range where the bicycle is stable can be maximized
within any physical actuator magnitude and bandwidth limits.
\citet{Schwab2008} elaborates on some of the possibilities in scheduling the
gains for such a controller and shows that a linear scheduling with respect to
speed can give satisfactory stability for a low speed range. We use this simple
feedback principle as the basis for our balance assist controller.

\section{Methods}
%
\subsection{Bicycle}
%
We modified an electric bicycle (Grenoble C8 HMB, Royal Dutch Gazelle, Dieren,
The Netherlands) with a custom motor in the steering assembly capable of
applying up to 7~\si{\newton\meter} of torque between the head tube and steer
tube, see Figure~\ref{fig:balance-assist-bicycle}. A custom motor controller
converts the commanded torque to applied torque. We measure the speed of the
rear wheel with an encoder (ABS Speed Sensor, Magura GmbH, Bad Urach, Germany)
and measure the body fixed roll rate of the bicycle with a MEMs rate gyroscope
(VR IMU BN0086, Sparkfun, Niwot, USA). The balance assist control algorithm is
implemented on a microprocessor (Teensy, PJRC, USA) and data from all sensors
is logged with a CAN bus (CanEdge2, CSS Electronics, Aabyhoej, Denmark) at at
least 100 Hz.

@@

Following the initial threshold determination, we choose perturbation forces
according to a random and adaptive staircase procedure applying perturbations
above and below the initial perturbation threshold, while allowing small
progression of the perturbation threshold to accommodate learning effects. Five
possible perturbation forces are determined based on the initial perturbation
threshold estimate: the initial estimate itself, two perturbations lower than
the initial estimate, and two perturbations higher than the initial estimate.
The five possible perturbations are separated by 10~\si{\newton} steps. For
example, if the initial estimate of the perturbation threshold of a participant
is 80~\si{\newton}, the five possible perturbations are 60, 70, 80, 90 and
100~\si{\newton}. Which one of these five perturbations is chosen is determined
at random. If the perturbation results in a fall, the estimate of the
perturbation threshold is decreased by 10~\si{\newton}, and vice versa if the
perturbation did not result in a fall. Five new possible perturbation forces
are determined around the updated perturbation threshold, and a new
perturbation is selected at random. This process iterates until twenty
perturbations are applied. The goal of this adaptive staircase procedure is to
have participants fall for approximately 50\% of the time. After the first set
of 20 perturbations, we let the cyclist rest and then perform another 20
perturbations. We randomize whether the balance assist system is on or off
during the first or second set of perturbations for each participant.

@@

We record the order number of each perturbation, i.e. first, second, third,
\ldots, to measure how many perturbations a rider is subjected to before the
current pertubration. At the initiation of each perturbation we log the
instantaneous steer and roll angles to characterize the configuration of the
bicycle when perturbed. The gain setting on the balance assist controller
indicates if the assistance is off \(g=0\) or on at two different levels: low
\(g=8\) or high \(g=10\). A recovery from the perturbation is successful if the
person neither places their foot down onto the treadmill surface nor their
wheel of the bicycle exits the width of the treadmill belt. We record this as a
binary variable \(f\) for "fall". All measured variables are reported in
Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall outcome'' and
    ``Perturbation order number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain & \(g\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} & wheel encoder \\
    Fall outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

\subsection{Statistics}
%
We test our hypothesis that the balance assistance controller will reduce the
probability of falling when perturbed externally at the handlebar. We have a
single binary fall outcome variable \(f\) that is dependent on several possible
explanatory independent variables, one of which is the binary balance
assistance (on or off). See Table~\ref{tab:stat-model-variables} for the
statistical model variables.
%
\begin{table}
  \centering
  \caption{Independent and dependent variables used in the statistical model.}
  \label{tab:stat-model-variables}
  \begin{tabular}{llll}
    \toprule
    Variable & Causality & Units & Description \\
    \midrule
    \(L\)        & independent & \si{\newton\meter\second} & angular impulse of perturbation torque \\
    \(\delta_0\) & independent & \si{\degree} & steer angle at start of perturbation \\
    \(\phi_0\)   & independent & \si{\degree} & roll angle at start of perturbation \\
    \(f\)        & dependent   & boolean & outcome: did not fall 0 or did fall 1 \\
    \(c\)        & independent & boolean & balance assist: off 0 or on 1 \\
    \(j\)        & independent & integer & order number of perturbation \\
    \bottomrule
  \end{tabular}
\end{table}

@@

Before fitting the model, we scale each independent variable \(x_{ij}^k\) such
that they have a mean of zero and a standard deviation of one by cluster-mean
centering, as recommended by \cite{Enders2007}, with clusters being an
individual subject. The clusters are chosen as all data from an individual
subject because we are only interested in the association between the state of
the balance assist system and the outcome of the perturbation. We expected
there to be a variation between participants in how well they are able to
resist the perturbation. However, this was not true. Cluster-mean centering
shows there to be no variation between participants. This fact allows us to
utilize a simple single-level logistic regression model, instead of a
multilevel model. This left us with angular impulse, perturbation order,
balance assist state, and roll \& steer angles at the time of perturbation as
independent variables. We also include interaction effects between the balance
assist state and the other four variables. We divide the analysis into two
separate model evaluations, one for the 6~\si{\kph}, \(g=10\) trials and one
for the 10~\si{\kph}, \(g=8\) trials and we evaluate our hypothesis for each
set of data. Statistics were computed with the lme4 version
1.1~\citep{Bates2015} package in R version 4.1.3~\citep{RCoreTeam2021}.

\section{Results}
%
The coefficient estimates for a single-level logistic regression for the data
from the 6~\si{\kph}, gain $g=10$ trials are shown in
Table~\ref{tab:freq-coefs-6}. The angular impulse, perturbation order, and
balance assist state are all statistically significant predictors with angular
impulse having a dominant effect. Larger angular impulse increases the
probability to fall and enduring more perturbations or having the balance
assist on, decrease the probability to fall. The associated multiplicative
change in odds are shown in Table~\ref{tab:freq-coefs-6}.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    6~\si{\kph} and gain $g=10$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-6}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept \(\beta\)                           & -0.29 & 0.17 & 0.09 & 0.75 & 0.53 & 1.05 \\
    Angular impulse \(L\)                         & 1.69 & 0.27 & 0.00\(^*\) & 5.40 & 3.18 & 9.16 \\
    Perturbation order \(j\)                      & -0.77 & 0.22 & 0.00\(^*\) & 0.46 & 0.30 & 0.72 \\
    Balance assist state \(c\)                    & -0.64 & 0.27 & 0.02\(^*\) & 0.53 & 0.31 & 0.89 \\
    Roll angle \(\phi_0\)                         & -0.25 & 0.21 & 0.24 & 0.78 & 0.51 & 1.18 \\
    Steer angle \(\delta_0\)                      & -0.14 & 0.21 & 0.51 & 0.87 & 0.58 & 1.32 \\
    Balance assist state $\times$ roll angle         & 0.52 & 0.34 & 0.12 & 1.68 & 0.86 & 3.29 \\
    Balance assist state $\times$ steer angle        & -0.41 & 0.34 & 0.22 & 0.66 & 0.34 & 1.28 \\
    Balance assist state $\times$ angular impulse    & 0.41 & 0.41 & 0.32 & 1.51 & 0.67 & 3.38 \\
    Balance assist state $\times$ perturbation order & -0.53 & 0.34 & 0.12 & 0.59 & 0.30 & 1.15 \\
    \bottomrule
  \end{tabular}
\end{table}

The coefficient estimates for a single-level logistic regression at
10~\si{\kph} with gain $g=8$ are shown in Table \ref{tab:freq-coefs-10}. The
angular impulse and perturbation order are statistically significant predictors
with angular impulse being the dominant effect. Larger angular impulse
increases the probability to fall and enduring more perturbations decreases the
probability to fall. Unlike the 6~\si{\kph} trials, the balance assist state is
not a significant predictor. At both speeds the angular impulse has about twice
the effect as the perturbation order.
%
\begin{table}
  \centering
  \caption{Logistic regression coefficient estimates \(\alpha_k\) at
    10~\si{\kph} and gain $g=8$ along with the standard error SE, p-value
    \(p\), multiplicative change in odds \(e^{\alpha_k}\), and the 5\%
    confidence interval bounds.}
  \label{tab:freq-coefs-10}
  \begin{tabular}{lrrrrrr}
    \toprule
    Variable & \(\alpha_k\) & SE & \(p\) & \(e^{\alpha_k}\) & 2.5\% & 97.5\% \\
    \midrule
    Intercept                & -0.24 & 0.16 & 0.13 & 0.78 & 0.57 & 1.07 \\
    Angular impulse \(L\)    & 2.39  & 0.29 & 0.00\(^*\) & 10.92 & 6.23 & 19.13 \\
    Perturbation order \(j\) & -1.16 & 0.21 & 0.00\(^*\)  & 0.31 & 0.21 & 0.48 \\
    Balance assist on \(c\)  & -0.44 & 0.24 & 0.07 & 0.64 & 0.40 & 1.03 \\
    Roll angle \(\phi_0\)    & 0.27  & 0.22 & 0.22 & 1.31 & 0.85 & 2.04 \\
    Steer angle \(\delta_0\) & -0.37 & 0.24 & 0.12 & 0.69 & 0.43 & 1.10 \\
    Balance assist state $\times$ roll angle         & -0.61 & 0.34 & 0.08 & 0.54 & 0.28 & 1.07 \\
    Balance assist state $\times$ steer angle        & 0.56  & 0.35 & 0.11 & 1.76 & 0.88 & 3.50 \\
    Balance assist state $\times$ angular impulse    & 0.46  & 0.44 & 0.29 & 1.59 & 0.68 & 3.74 \\
    Balance assist state $\times$ perturbation order & -0.37 & 0.32 & 0.25 & 0.69 & 0.37 & 1.30 \\
    \bottomrule
  \end{tabular}
\end{table}

Turning the balance assist system on significantly \((p<0.05)\) reduces the
odds that a perturbation results in a fall while cycling at a speed of
6~\si{\kph}. Figure~\ref{fig:probability-6kph} visualizes the probability of
falling as a function of the mean and centered angular impulse per participant
for the balance assist state on and off while keeping all other explanatory
factors at their centered mean values. This figure is created by setting all
explanatory variables to there mean and calculating the probability from
Eq~\ref{eq:log-regress} for only change in angular impulse given the estimates
in Table~\ref{tab:freq-coefs-6}. The table indicates that the balance assist
system halves (0.53) the odds that a perturbation results in a fall. This
figure shows that for relatively large impulses the probability to fall is
unity for both states of balance assist on and off. And for relatively small
impulses the probability to fall is null for both states. But for impulses in
the magnitude region (-1 to 0.5 STD), i.e. around the mean-centered fall
threshold, the probability of falling is significantly lowered with the balance
assist system on. The skewness of the probability curves arrives from the
interaction effects.\todo{check this statement about skewedness this and think
about it} Figure~\ref{fig:probability-10kph} shows the same result for the
10~\si{\kph} trials which has a similar trend of reducing the probability to
fall with the balance assist system turned on, but the effect is not
significant.
%
\begin{figure}
  \centering
  % NOTE : We could uncomment the longer subcaption if titles removed from
  % figures. It is currently redundant.
  \subcaptionbox{Fall probability for the 6~\si{\kph} trials.
    \label{fig:probability-6kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_6kmh.png}
  }
  \subcaptionbox{Fall probability for the 10~\si{\kph} trials.
    \label{fig:probability-10kph}}{
    \includegraphics[width=80mm]{figures/predicted_fall_probability_10kmh.png}
  }
  \caption{Comparison of predicted fall probability for balance assist on
    (solid) or off (dashed) when all other predictor variables are fixed at zero,
    except for the interaction effect between the balance assist and the
    angular impulse. The abscissa is standard deviation around the mean (all
    perturbations), centered per participant.}
  \label{fig:probability}
\end{figure}

\section{Discussion}
%
We have shown that at a 6~\si{\kph} riding speed the addition of balance assist
control reduces the chance that a rider will fall when perturbed around the
limits of their control authority. But this effect diminishes just below
significance at the higher speed scenario of 10~\si{\kph}. We were only able to
test these two speed-gain scenarios for mostly homogeneous sets of riders
within the resources of this research project, but additional experimental work
could help understand more completely the range and limits of the positive
effect of the balance system. For example, it is possible that simply
increasing the controller gain at 10~\si{\kph} also results in a significant
positive effect. A longitudinal study of normal use of the balance assist
bicycle compared to a control group could provide the strongest evidence of any
benefit we have seen in this more narrow scenario.
